---
title: 'occOutlierR: Identifying outlying species occurrence records'
author: "Cory Merow, Gonzalo E. Pinilla-Buitrago"
date: "2024-09-09"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Title of Your Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, eval = TRUE,
                      fig.width = 4, fig.height = 4, dpi = 200, out.width = '80%')
```

## Outlier Detection in Species Distribution Modeling

Species distribution models (SDMs) are widely used in ecology to predict the potential distribution of species across different environments. However, building accurate SDMs can be challenging due to the presence of outliers in occurrence data. Outliers are records that deviate significantly from other observations and can have a substantial impact on model predictions.

The difficulty of detecting outliers lies in their unpredictability. They can arise from measurement errors, sampling processes, or skewed distributions. Small sample sizes exacerbate this issue, making it challenging to identify outliers using traditional descriptive statistics such as boxplots and percentiles. Furthermore, the presence of disjunct clusters of outlying records (known as "masked outliers") can make detection even more difficult.

In this vignette, we will show several approaches for identifying and dealing with these issues by examining various methods for detecting outliers spatial and environmental outliers. The  goal is to provide a comprehensive overview on how to effectively address these issues using adaptable outlier detection methods.

```{r libraries}
library(occOutliers)
```

## Detection of outliers
Outliers are identified based on their distance from other observations in either geographic or environmental space. A multivariate distance metrics, such as Euclidean distance, are used to measure the distance between each record and the centroid of all observations. One-tailed tests for outlier detection are designed to detect exceedingly large distances which are practical for automation in modeling pipelines.

### Spatial
To calculate spatial outliers, we need to use a method that takes into account the spatial relationships between observations. This can be achieved by calculating the distances between each observation and the centroid of all observations. We require an `sf` object with a defined Coordinate Reference System (CRS) for this calculation. If the CRS is projected, we will use Euclidean distance as a measure of distance. However, if the CRS is geographic, we will use Great Circle distance to account for the curvature of the Earth. The `occOutlier` package uses `sf::st_distance()` internally to perform these calculations, which provides a convenient interface for working with spatial data in R.

In these vignette's examples, we will use species geographic coordinates (WGS84). Each CSV file contain the coordinates of each observation ('X' and 'Y'), plus more columns with environmental data which will be use in upcoming examples. For this first example, we will use *Astraea macroura* to detect spatial outliers.

```{r}
a_macroura <- read.csv(system.file('extdata/Astraea_macroura.csv',
                                   package = 'occOutliers')) |>
  sf::st_as_sf(coords = c("X", "Y")) |>
  # Set WGS84
  sf::st_set_crs(4326)
plot(a_macroura[1])
```

´occOutlier´ offers four methods to detect outlier based on their distance to the centroid.

- 'iqr': Interquartile range
- 'grubbs': The Grubb’s outlier test
- 'dixon': The Dixon's outlier test
- 'rosner': The Rosner test for multiple outliers

#### Interquartile range (IQR)
The IQR method calculates the difference between the 75th percentile (Q3) and the 25th percentile (Q1). Data points more than 1.5 times this range away from Q1 or Q3 are typically considered outliers. This simple yet effective approach helps identify data points that significantly deviate from the rest of the dataset, making it a useful tool in outlier detection.

The main function of the package is `findOutliers()`. The main input is a `sf` object with coordinates. Then, you can define if you want to detect spatial or environmental outliers with their respective parameters (both TRUE as default). And finally you can define the method to detect outliers with the 'method' parameter. Here, we will use 'iqr'.

```{r sp_iqr}
sp_iqr <- findOutliers(a_macroura, environmental = FALSE, method = "iqr")
```
As you notice for the message, two outliers were detected using this method. The output of the function is the same object with an additional column (out_spatial), that allow you to know which are the outliers.

```{r sp_iqr_table}
sp_iqr$out_spatial
```
To visualize, you can plot the points using plot() function.
```{r}
plot(sp_iqr["out_spatial"])
```

This package also provide the `plotOutlier` function to easy visualize the result by adding a shapefile.
```{r plotOut}
# Load world shapefile
worldWGS84 <- readRDS(system.file('extdata/worldWGS84.rds',
                                  package = 'occOutliers'))
# Get outlier plot
plotOutliers(sp_iqr, shpToPlot = worldWGS84)
```

As a tip, keep in mind that the plotOutlier() function has two useful parameters that allow you to expand the shapefile buffer around coordinates (bufferDist), and change the legend position (legLoc).

```{r plotOutMore}
plotOutliers(sp_iqr, legLoc = "bottomright", bufferDist = 0.4, 
             shpToPlot = worldWGS84)
```

#### The Grubb’s outlier test
The Grubbs test is a statistical method used to detect outliers in a dataset. It calculates the distance between each data point and the mean of the remaining data points, then determines which data points are farthest away from the mean. This test is useful when there are multiple outliers in the dataset or when the data distribution is skewed. This test is the default method for `findOutlier()`.

```{r sp_grubbs}
sp_grubbs <- findOutliers(a_macroura, environmental = FALSE)
```
Here, it was just detected one spatial outlier.
```{r sp_grubbs_plot}
plotOutliers(sp_grubbs, shpToPlot = worldWGS84)
```

#### The Dixon's outlier test
Dixon's test is a statistical method used to detect single outliers in a dataset. It calculates the difference between the largest and smallest values, then compares this range to a critical value from a standard normal distribution. This test assumes that there are no multiple outliers, so it may not be suitable for datasets with complex outlier patterns.

```{r sp_dixon}
sp_dixon <- findOutliers(a_macroura, method = "dixon", environmental = FALSE)
```
```{r sp_dixon_plot}
plotOutliers(sp_dixon, shpToPlot = worldWGS84)
```

#### Rosner's Test
Rosner's test is a statistical method that detects single or multiple outliers in a dataset by calculating the modified Z-score of each data point. It provides an effective way to identify anomalies in datasets with large sample sizes or complex distributions, while being robust against non-normality and heavy-tailed distributions. This test requires that you define the number of outliers to test (parameter 'kRosner')

```{r sp_rosner}
sp_rosner <- findOutliers(a_macroura, method = "rosner", kRosner = 2,
                          environmental = FALSE)
```
```{r sp_rosner_plot}
plotOutliers(sp_rosner, shpToPlot = worldWGS84)
```

Keep in mind that you also can change the *p-value* for assessing the significant of the test for 'grubbs', 'dixon', and 'rosner'. Default is 1e-05. If we change to a smaller pval in the previous test, no outliers will be detected.
```{r sp_rosner_pval}
sp_pval <- findOutliers(a_macroura, method = "rosner", kRosner = 2,
                          environmental = FALSE, pval = 1e-06)
plotOutliers(sp_pval, shpToPlot = worldWGS84)
```

### Environmental outliers
Similar to the process of finding spatial outliers, you can use the same four methods to find environmental outliers. However, for this process you will also be able to select several ways to measure the distance between the centroid and the observations ('distEnvMethod' parameter), and if you which to scale data ('scaleData'; default is TRUE). The output will be an `sf` object with a new column 'out_env'.

For this example, we will use *Anarthria scabra* using Scrubb´s test but changing the method to measure distance bewteen centroids.
```{r A_scraba}
A_scraba <- read.csv(system.file('extdata/Anarthria_scabra.csv',
                                   package = 'occOutliers')) |>
  sf::st_as_sf(coords = c("X", "Y")) |>
  # Set WGS84
  sf::st_set_crs(4326)
plot(A_scraba[1])
```

#### Euclidean distance
The Euclidean distance is a measure of the straight-line distance between two points in n-dimensional space, calculated as the square root of the sum of squared differences between corresponding coordinates.
```{r e_euc}
e_euclidean <- findOutliers(A_scraba, spatial = FALSE)
```
Here, one environmental outlier is detected, which correpons to the northern observation.
```{r e_euc_plot}
plotOutliers(e_euclidean, shpToPlot = worldWGS84, legLoc = "topright")
```

#### Manhattan distance
The Manhattan distance (also known as L1 distance) is a measure of the sum of absolute differences between corresponding coordinates of two points in n-dimensional space. It's named after the grid layout of New York City streets.
```{r e_man}
e_manhattan <- findOutliers(A_scraba, spatial = FALSE, 
                            distEnvMethod = "manhattan")
```
```{r e_man_plot}
plotOutliers(e_manhattan, shpToPlot = worldWGS84, legLoc = "topright")
```

#### Cosine
The Cosine Dissimilarity measures the similarity between two vectors by calculating the cosine of the angle between them, resulting in a value between -1 and 1, where higher values indicate more similar vectors.
```{r e_cos}
e_cosine <- findOutliers(A_scraba, spatial = FALSE, 
                         distEnvMethod = "cosine")
```
As the Manhattan distance example, it was not detected an environmental outlier.

#### Mahalanobis
The Mahalanobis distance is a measure of the distance between a point and a distribution, taking into account the covariance between variables. It's used to detect outliers in multivariate data by calculating how far a point is from the center of the distribution while accounting for its spread and correlations.

```{r e_mah}
e_mahalanobis <- findOutliers(A_scraba, spatial = FALSE, 
                              distEnvMethod = "mahalanobis")
```
```{r}
plotOutliers(e_mahalanobis, shpToPlot = worldWGS84, legLoc = "topright")
```

### Spatial & Environmental
You can also detect both types of outliers in an unique call. Here, we will use *Camissonia_tanacetifolia* as an example.
```{r C_tan}
C_tanacetifolia <- read.csv(system.file('extdata/Camissonia_tanacetifolia.csv',
                                   package = 'occOutliers')) |>
  sf::st_as_sf(coords = c("X", "Y")) |>
  # Set WGS84
  sf::st_set_crs(4326)
out_C_tana <- findOutliers(C_tanacetifolia)
```
In this example, it was detected two outliers, one spatial and one environmental, but they correspond to the same observation.

```{r C_tan_plot}
plotOutliers(out_C_tana, shpToPlot = worldWGS84, legLoc = NULL)
```

Changing the p-value to 0.05 (a lower value), you are able to detect four outliers. One exclusively spatial, one exclusively environmental, and two other that are outliers in the geography and in the environment.

```{r out_pval}
out_pval <- findOutliers(C_tanacetifolia, pval = 0.05)
plotOutliers(out_pval, shpToPlot = worldWGS84, legLoc = NULL)
```

<!-- In this case, because there are a cluster of presences in North America, using a lower p-value helps to detect them.  -->

<!-- ```{r,fig.width=7, fig.height=7, dpi=300,out.width='80%'} -->
<!-- myPres4=read.csv(system.file('extdata/SpeciesCSVs/Aster_tataricus.csv',package='occOutliers')) -->
<!-- myPres4=myPres4[complete.cases(myPres4),] -->
<!-- sp::coordinates(myPres4)=c(1,2) -->
<!-- myPres4DF=sp::SpatialPointsDataFrame(myPres4,data.frame(raster::extract(myEnv,myPres4))) -->
<!-- presOut4=findOutlyingPoints(pres=myPres4DF, -->
<!--                            spOutliers=TRUE, -->
<!--                            envOutliers=TRUE, -->
<!--                            pval=1e-3) # note the lower value here -->
<!-- plotOutliers(presOut4,shpToPlot = world.shp) -->
<!-- ``` -->
<!-- However, lowering the p-value cannot just be done arbitarily. In this case, the biased sampling in Europe places the geographic centroid there, making the more poorly sampled parts of Asia appear as outliers. Hence some visual inspection should be used to assess results; this issue occurs typically when extreme sampling bias is present. A lower p-value of 1e-6 helps this situation, although it's unclear whether the detected outliers still reflect undersampling. If one suspects that they do, it would be preferrable to spatially thin the presences (e.g., using the `spThin` package) to remove some sampling bias before using outlier detection.  -->

<!-- ```{r,fig.width=6, fig.height=6, dpi=300,out.width='80%'} -->
<!-- myPres5=read.csv(system.file('extdata/SpeciesCSVs/Betula_pendula.csv',package='occOutliers')) -->
<!-- myPres5=myPres5[complete.cases(myPres5),] -->
<!-- sp::coordinates(myPres5)=c(1,2) -->
<!-- myPres5DF=sp::SpatialPointsDataFrame(myPres5,data.frame(raster::extract(myEnv,myPres5))) -->
<!-- presOut5=findOutlyingPoints(pres=myPres5DF, -->
<!--                            spOutliers=TRUE, -->
<!--                            envOutliers=TRUE, -->
<!--                            pval=1e-3) # note the lower value here -->
<!-- plotOutliers(presOut5,shpToPlot = world.shp) -->
<!-- presOut5.1=findOutlyingPoints(pres=myPres5DF, -->
<!--                            spOutliers=TRUE, -->
<!--                            envOutliers=TRUE, -->
<!--                            pval=1e-6) # note the lower value here -->
<!-- plotOutliers(presOut5.1,shpToPlot = world.shp) -->
<!-- ``` -->